First and foremost, I would like to express my gratitude towards my wonderful supervisors \textbf{Dr~Catherine~Watson} and \textbf{Dr~Richard~Clarke} for their endless support. I truly appreciated the calm but firm guidance you both gave me as I navigated this learning experience. Your patience, expertise and good humour definitely defined this project, and I felt like I was in good hands from the first day. Also, the philosophical tangents during our meetings were always very enjoyable. 

Catherine, thank you for your introducing me to the field of acoustic phonetics, a wonderful bridge between linguistics and computer science. Your previous work and experience in the field formed key foundations for my project, and the way you shared your passion for the research was infectious. To Richard, your advice on how to best communicate this research in the context of a Part IV Project was truly invaluable. I found your reassuring yet realistic perspective of these projects very useful as I prepared the products of my research.

A big thanks to \textbf{Xiao Bo Lu} and \textbf{Peter Bier} who, alongside Dr Catherine Watson, helped to develop the method and write the Perl, MATLAB and CMGUI scripts used in this project for extracting phonetic information from MRI images. This image processing was the foundation of all the analyses in this project. I would also like to thank \textbf{Daniel Tan}, \textbf{Kalyan Chilukuri} and \textbf{Helen Searle} whose previous undergraduate and postgraduate research work contributed to the analysis of this data set and the refinement of this system of MRI image processing. I also thank Dr Catherine Watson again for her work in gathering this MRI data set, under the ethics approval granted by The University of Auckland Human Participants Ethics Committee (Reference number 2010/136, granted 12/05/2010).

I would also like to thank \textbf{Dr~Richard~Christie} and doctoral candidate \textbf{Stephen~Waite} for assisting me in using CMGUI during the initial data collection phase of this project.

Lastly, I would like to thank \textbf{The~Department~of Engineering~Science} and all its members over the last five years. You truly make this department the melting pot of friendship and learning that it is.